---
title: "Tramacastilla_disturbance_palaeoecology_figures"
author: "Irene Julián-Posada"
date: "June 2026"
output: pdf_document
---
This is the code used to obtain figures and tables included in Julián-Posada et al. (in prep).

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
packages <- c("readr", "tidyverse", "patchwork", "lavaan", #for data manipulation and visualisation
              "semPlot", #for structural equation modelling visualisation
              "rioja") #spore and pollen diagrams
lapply(packages, library, character.only = TRUE)

load("tramacastilla_disturbance_palaeoecology.RData")
```

# 1. Table 1 - Results summary

Pollen data
```{r}
pollen_summary <- data.frame(
  Proxy = "Pollen",
  Metric = "Number of pollen grains per sample",
  Min = min(tram20_pollen_holocene$`Terrestrial pollen sum`),
  Max = max(tram20_pollen_holocene$`Terrestrial pollen sum`),
  Mean = round(mean(tram20_pollen_holocene$`Terrestrial pollen sum`), 2),
  SD = round(sd(tram20_pollen_holocene$`Terrestrial pollen sum`), 2)) %>%
  rbind(data.frame(
    Proxy = "Pollen",
    Metric = "Pollen richness: number of pollen types per sample",
    Min = min(tram20_pollen_holocene$`No of taxa`),
    Max = max(tram20_pollen_holocene$`No of taxa`),
    Mean = round(mean(tram20_pollen_holocene$`No of taxa`), 2),
    SD = round(sd(tram20_pollen_holocene$`No of taxa`), 2))) %>%
  rbind(data.frame(
    Proxy = "Pollen",
    Metric = "PAR (number of grains·cm-2·yr-1)",
    Min = round(min(tram20_PAR_sum$sum), 2),
    Max = round(max(tram20_PAR_sum$sum), 2),
    Mean = round(mean(tram20_PAR_sum$sum), 2),
    SD = round(sd(tram20_PAR_sum$sum), 2)))
```

Fungal spores' data
```{r}
fungal_spores_summary <- data.frame(
  Proxy = "Fungal spores'",
  Metric = "Number of fungal spores per sample",
  Min = min(tram20_spores$total_spores),
  Max = max(tram20_spores$total_spores),
  Mean = round(mean(tram20_spores$total_spores), 2),
  SD = round(sd(tram20_spores$total_spores), 2)) %>%
  rbind(data.frame(
    Proxy = "Fungal spores'",
    Metric = "Fungal richness: number of fungal types per sample",
    Min = min(tram20_spores$Noftaxa),
    Max = max(tram20_spores$Noftaxa),
    Mean = round(mean(tram20_spores$Noftaxa), 2),
    SD = round(sd(tram20_spores$Noftaxa), 2))) %>%
  rbind(data.frame(
    Proxy = "Fungal spores'",
    Metric = "SAR (number of spores·cm-2·yr-1)",
    Min = round(min(tram20_SAR$total_SAR), 2),
    Max = round(max(tram20_SAR$total_SAR), 2),
    Mean = round(mean(tram20_SAR$total_SAR), 2),
    SD = round(sd(tram20_SAR$total_SAR), 2)))
```

Charcoal data
```{r}
charcoal_summary <- data.frame(
  Proxy = "Charcoal",
  Metric = "Number of charcoal particles per sample",
  Min = min(CharTotS.110.age$CharCount),
  Max = max(CharTotS.110.age$CharCount),
  Mean = round(mean(CharTotS.110.age$CharCount), 2),
  SD = round(sd(CharTotS.110.age$CharCount), 2))  %>%
  rbind(data.frame(
    Proxy = "Charcoal",
    Metric = "Charcoal area per sample (mm2)",
    Min = min(CHAR$sumCharArea)/1e6,
    Max = round(max(CHAR$sumCharArea)/1e6, 2),
    Mean = round(mean(CHAR$sumCharArea/1e6), 2),
    SD = round(sd(CHAR$sumCharArea/1e6), 2))) %>%
  rbind (data.frame(
    Proxy = "Charcoal",
    Metric = "Width to length (W/L) ratio",
    Min = round(min(seedles$NPath.WLRatio), 2),
    Max = round(max(seedles$NPath.WLRatio), 2),
    Mean = round(mean(seedles$NPath.WLRatio), 2),
    SD = round(sd(seedles$NPath.WLRatio), 2))) %>%
  rbind (data.frame(
    Proxy = "Charcoal",
    Metric = "CHAR (µm2·cm-2·yr-1)",
    Min = min(CHAR$CHAR.tram),
    Max = max(CHAR$CHAR.tram),
    Mean = round(mean(CHAR$CHAR.tram), 2),
    SD = round(sd(CHAR$CHAR.tram), 2)))
```

SedaDNA data
```{r}
RAI_plants_sum <- RAI_plants %>%
  mutate(total_RAI_plants = rowSums(across(`Abies alba`:Ziziphora))) %>%
  mutate(richness_plants = rowSums(select(., `Abies alba`:Ziziphora) > 0, na.rm = TRUE)) %>%
  select(total_RAI_plants, richness_plants)

RAI_animals_sum <- RAI_animals %>%
  mutate(total_RAI_animals = rowSums(across(`Bos taurus`:`Nyctalus noctula`))) %>%
  mutate(richness_animals = rowSums(select(., `Bos taurus`:`Nyctalus noctula`) > 0, na.rm = TRUE)) %>%
  select(total_RAI_animals, richness_animals)

sedaDNA_summary <- data.frame(
  Proxy = "sedaDNA",
  Metric = "Plant taxa per sample",
  Min = min(RAI_plants_sum$richness_plants),
  Max = max(RAI_plants_sum$richness_plants),
  Mean = round(mean(RAI_plants_sum$richness_plants), 2),
  SD = round(sd(RAI_plants_sum$richness_plants), 2)) %>%
  rbind (data.frame(
    Proxy = "sedaDNA",
    Metric = "Animal taxa per sample",
    Min = min(RAI_animals_sum$richness_animals),
    Max = max(RAI_animals_sum$richness_animals),
    Mean = round(mean(RAI_animals_sum$richness_animals), 2),
    SD = round(sd(RAI_animals_sum$richness_animals), 2))) %>%
  rbind (data.frame(
    Proxy = "sedaDNA",
    Metric = "Plant RAI",
    Min = round(min(RAI_plants_sum$total_RAI_plants), 2),
    Max = round(max(RAI_plants_sum$total_RAI_plants), 2),
    Mean = round(mean(RAI_plants_sum$total_RAI_plants), 2),
    SD = round(sd(RAI_plants_sum$total_RAI_plants), 2))) %>%
  rbind (data.frame(
    Proxy = "sedaDNA",
    Metric = "Animal RAI",
    Min = round(min(RAI_animals_sum$total_RAI_animals), 2),
    Max = round(max(RAI_animals_sum$total_RAI_animals), 2),
    Mean = round(mean(RAI_animals_sum$total_RAI_animals), 2),
    SD = round(sd(RAI_animals_sum$total_RAI_animals), 2)))
```

Create table 1
```{r}
table1 <- rbind(pollen_summary, fungal_spores_summary, charcoal_summary, sedaDNA_summary)
table1
```

# 2. Figure 2 - sedimentation rate + spores + animals

Include all data in one dataframe
```{r}
sed.rate.SAR.diagram <- tram20_SAR %>%
  dplyr::select(-c(depth, total_SAR)) %>%
  left_join(tram20_depth_ages_sed_rate %>%
              dplyr::select(age, sed_rate), by = "age") %>%
  relocate(sed_rate, .after = "age") %>%
  relocate(`Obligate coprophilous`, .after = "Facultative coprophilous") %>%
  full_join(RAI_animals_domestic, by = "age") %>%
  arrange(age) %>%
  filter(age<12000) %>%
  mutate(sed_rate = sed_rate*1000, # CAREFUL WITH THIS TRANSFORMATION -- just for visualisation
         Cattle = Cattle*100, # CAREFUL WITH THIS TRANSFORMATION -- just for visualisation
         Sheep = Sheep*100, # CAREFUL WITH THIS TRANSFORMATION -- just for visualisation
         Goat = Goat*100) # CAREFUL WITH THIS TRANSFORMATION -- just for visualisation
```

Configuration of styles
```{r}
ex.spores <- c(FALSE, rep(TRUE, times = 28), rep(FALSE, times = 3))
c.sed.rate.line <- c("black", rep(NA, times = 31))
c.spores.bar <- c(NA, rep("black", times = 28), NA, NA, NA)
c.animals.bar <- c(rep(NA, times = 29), rep("#E57E7E", times = 3))
c.spores.poly <- c(NA, rep("#458B74", times = 14), rep("#993F00", times = 4),
                   rep("orange", times = 8), rep("#458B74", times = 1), rep("#993F00", times = 1), NA, NA, NA)
```

Create figure
```{r}
p.sedrate.sar <- strat.plot(sed.rate.SAR.diagram %>%
                              dplyr::select(-age) %>%
                              rename("sedimentation \n rate (cm/yrs)" = "sed_rate"),
                            yvar = sed.rate.SAR.diagram %>%
                              dplyr::select(age)/1000,
                            ylabel = "Age (ka BP)", y.tks= seq(1, 12, 1), y.rev=TRUE,
                            plot.line=TRUE, plot.poly=FALSE, plot.bar=FALSE, col.line=c.sed.rate.line,
                            scale.percent=TRUE, xSpace=0.009, x.pc.inc=10, x.pc.lab=TRUE, x.pc.omit0=TRUE, las=2, srt.xlabel=70, exag = FALSE)
p.sedrate.sar <- strat.plot(sed.rate.SAR.diagram %>%
                              dplyr::select(-age),
                            yvar = sed.rate.SAR.diagram %>%
                              dplyr::select(age)/1000, y.rev=TRUE,
                            plot.line=TRUE, plot.poly=TRUE, plot.bar=TRUE, col.line=c.spores.bar, col.bar=c.spores.bar, col.poly=c.spores.poly,
                            scale.percent=TRUE, xSpace=0.009, x.pc.inc=10, x.pc.lab=FALSE, x.pc.omit0=TRUE, las=2, exag = ex.spores, col.exag = "auto",
                            x.axis=FALSE, x.names = NA, y.pc.lab = FALSE, add=TRUE)
p.sedrate.sar <- strat.plot(sed.rate.SAR.diagram %>%
                              dplyr::select(-age),
                            yvar = sed.rate.SAR.diagram %>%
                              dplyr::select(age)/1000, y.rev=TRUE,
                            plot.line=FALSE, plot.poly=FALSE, plot.bar=TRUE, col.bar= c.animals.bar, lwd.bar = 2,
                            scale.percent=TRUE, xSpace=0.009, x.pc.inc=10, x.pc.lab=FALSE, x.pc.omit0=TRUE, las=2,
                            x.axis=FALSE, x.names = NA, y.pc.lab = FALSE, add=TRUE) #save pdf A4 and MODIFY SCALES
```

# 3. Figure 3 - Changes in disturbance regimes with RoC, cpt and fire events

Figure 3a - Fire events
```{r}
p.fig3.1 <- tr_glob_exp %>%
  ggplot(aes(x = age_top_i/1000, y = acc_i)) +
  geom_col(width = 0.01, fill = "grey40") +
  geom_smooth(method = "loess", span = 0.15, se = FALSE, color = "red", size = 0.5) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, color = "grey", size = 0.5) +
  geom_point(data = tr_glob_exp %>% filter(peaks_pos_sig > 0),
             aes(x = age_top_i/1000, y = peaks_pos_sig*30000),
             shape = 3, size = 2, color = "red") +
  geom_col(data = tr_glob_exp %>% filter(peaks_pos_sig > 0),
           aes(x = age_top_i/1000, y = peaks_pos_sig*30000),
           fill = "red", alpha = 0.3, width = 0.02) +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  scale_y_continuous(breaks = c(0, 10000, 20000),
                     labels = c("0", "1e+4", "2e+4")) +
  theme_classic() +
  ylab("") +
  theme(axis.title.x = element_blank(),
        legend.position = "right",
        plot.margin = margin(0, 0, 0, 0, "cm"),
        legend.text = element_text(size = 11))
```

Figure 3b - W/L ratio
```{r}
p.fig3.2 <- W.L.ratio %>%
  rename("Type of fuel" = "fuel") %>%
  ggplot(aes(x = age/1000, y = count, fill = `Type of fuel`, color = `Type of fuel`)) +
  geom_area(alpha = 0.6, position = "stack") +
  geom_smooth(method = "loess", se = FALSE, aes(color = `Type of fuel`), linewidth = 1, alpha = 0.8) +
  scale_y_continuous(trans = "sqrt",  # transformation with squared root
                     breaks = c(0, 5, 20, 50)) +
  labs(y = "") +
  theme_classic() +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  scale_fill_manual(values = c("Woody" = "darkgreen", "Grasslands" = "#FFD700")) +
  scale_color_manual(values = c("Woody" = "darkgreen", "Grasslands" = "#FFD700")) +
  theme(axis.title.x = element_blank(),
        legend.position = "right",
        plot.margin = margin(0, 0, 0, 0, "cm"),
        legend.text = element_text(size = 11)) +
  annotate("text", x = 2, y = 120, label = "peak = 59", vjust = -1, color = "red")
```

Figure 3c - CHAR with changepoints
```{r}
p.fig3.3 <- tram20_CHAR_cpt %>%
  ggplot(aes(x = age/1000, y = CHAR.tram)) +
  geom_line() +
  geom_vline(xintercept = changepoint_ages_CHAR/1000, color = "red", linetype = "dashed") +
  theme_classic() +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  scale_y_continuous(breaks = c(0, 10000, 20000, 30000),
                     labels = c("0", "1e+4", "2e+4", "3e+4")) +
  ylab("CHAR (µm2·cm-2·yr-1)") +
  theme(axis.title.x = element_blank())
```

Figure 3d - Rate of Change of SAR
```{r}
p.fig3.4 <- scenario_4_SAR_peak %>%
  ggplot(aes(x = Age/1000))+
  theme_classic() +
  geom_ribbon(aes(ymin = ROC_dw, ymax = ROC_up), fill = "grey") +
  geom_line(aes(y = ROC)) + 
  geom_point(size = 3, aes(y = ROC_peak), shape = 21, fill = "green") +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  ylab("RoC of SAR") +
  theme(axis.title.x = element_blank())
```

Figure 3e - Rate of Change of PAR
```{r}
p.fig3.5 <- scenario_4_PAR_peak %>%
  ggplot(aes(x = Age/1000))+
  theme_classic() +
  geom_ribbon(aes(ymin = ROC_dw, ymax = ROC_up), fill = "grey") +
  geom_line(aes(y = ROC)) + 
  geom_point(size = 3, aes(y = ROC_peak), shape = 21, fill = "green") +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  labs(y = "RoC of PAR") +
  theme(axis.title.x = element_blank())
```

Figure 3f - Rate of Change of RAI
```{r}
p.fig3.6 <- scenario_4_RAI_peak %>%
  ggplot(aes(x = Age/1000))+
  theme_classic() +
  geom_ribbon(aes(ymin = ROC_dw, ymax = ROC_up), fill = "grey") +
  geom_line(aes(y = ROC)) + 
  geom_point(size = 3, aes(y = ROC_peak), shape = 21, fill = "green") +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  labs(x = "Age (ka BP)", y = "RoC of RAI")
```

Create figure 3
```{r}
p.fig3.1/p.fig3.2/p.fig3.3/p.fig3.4/p.fig3.5/p.fig3.6 #save A5
```

# 4. Table 2 - Generalized Additive Models (GAMs)
``` {r}
extract <- function(summary_gam, var_name, predictor_name) {
  coef <- summary_gam$p.table[2,]
  data.frame(Variable=var_name, Predictor=predictor_name, 
             Estimate=coef[1], `Standard error`=coef[2], `t-value`=coef[3], 
             `p-value`=ifelse(coef[4]<0.001,"< 0.001 **",paste(round(coef[4],3),"**")),
             Deviance=paste0(round(summary_gam$dev.expl*100),"%"))
}

rbind(
  extract(summary_gam1, "Rate of Change sedaDNA", "Temperature"),
  extract(summary_gam2, "Rate of Change sedaDNA", "Cattle"),
  extract(summary_gam3, "Rate of Change sedaDNA", "Sheep"),
  extract(summary_gam4, "Rate of Change sedaDNA", "Obligate coprophilous fungi"),
  extract(summary_gam5, "Rate of Change sedaDNA", "Fire"),
  extract(summary_gam1_PAR, "Rate of Change pollen", "Temperature"),
  extract(summary_gam2_PAR, "Rate of Change pollen", "Cattle"),
  extract(summary_gam3_PAR, "Rate of Change pollen", "Sheep"),
  extract(summary_gam4_PAR, "Rate of Change pollen", "Obligate coprophilous fungi"),
  extract(summary_gam5_PAR, "Rate of Change pollen", "Fire"))
```

# 5. Table 3 - Parameters of Structural Equation Model (SEM)
``` {r}
parameterEstimates(fit_no_normal_8, standardized = TRUE) %>%
  filter(op == "~") %>%
  dplyr::select(lhs, rhs, est, se, z, std.all, pvalue) %>%
  rename("Response" = "lhs",
         "Predictor" = "rhs",
         "Estimate" = "est",
         "Standard error" = "se",
         "z-value" = "z",
         "Standardised estimate" = "std.all",
         "p-value" = "pvalue") 
```

# 6. Figure 4 - Structural Equation Model (SEM)
``` {r}
semPaths(fit_no_normal_8,
         what = "std",
         whatLabels = "std",
         style = "lisrel",
         layout = "tree",
         nodeLabels = c("Cattle", "Sheep", "Obligate coprophilous fungi", "Abies-Fagus",
                        "Temperate forests", "Evenness", "Grasslands", "Pinewoods", "Mediterranean", "d13C Mendukilo", "CHAR",
                        "Herbivory"),
         edge.label.cex = 1.2,
         sizeMan = 8,
         sizeLat = 8,
         label.cex = 2,
         mar = c(8, 1, 6, 1))
```

# 7. Figure 5 - Cross-correlations

Cross correlations before 6.5 ka BP
``` {r}
fig_cross_before <- p.richness.ccf_before/p.evenness.ccf_before/p.poaceae.ccf_before/
  p.pinus.ccf_before/p.betula.ccf_before/p.corylus.ccf_before/p.decquercus.ccf_before +
  plot_layout(ncol = 1, nrow = 9) +
  theme(plot.margin = margin(0, 0, 0, 0, "pt")) +
  plot_annotation(title = "Before 6.5 ka BP")
```

Cross correlations after 6.5 ka BP
``` {r}
fig_cross_after <- p.richness.ccf_after/p.evenness.ccf_after/ p.poaceae.ccf_after/p.pinus.ccf_after/
  p.betula.ccf_after/p.corylus.ccf_after/p.decquercus.ccf_after/p.abies.ccf_after/p.fagus.ccf_after +
  plot_layout(ncol = 1, nrow = 9) +
  theme(plot.margin = margin(0, 0, 0, 0, "pt")) +
  plot_annotation(title = "After 6.5 ka BP")
```

Cross correlations before and after 6.5 ka BP
``` {r}
fig_cross_before|fig_cross_after +
  plot_layout(guides = "collect")
```