---
title: "Tramacastilla_disturbance_palaeoecology_analyses"
author: "Irene Julián-Posada"
date: "June 2026"
output: pdf_document
---

This is the code used to perform statistical analyses included in Julián-Posada et al. (in prep).

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
packages <- c("readr", "tidyverse", "patchwork", "ggrepel", #for data manipulation and visualisation
              "strucchange", #breakpoint analysis
              "vegan", #rarefaction and eveness
              "Bchron", "RRatepol", #rate of change - RRatepol only works with R 4.2.0
              "EnvCpt", "segmented", #changepoint analysis
              "lavaan", "semPlot", "corrplot", #structural equation modelling
              "TSA", #cross-correlations
              "FactoMineR", "factoextra", "arm", "psych", #multivariate analyses and multiple factor analysis
              "readxl", #read excel files
              "nlme", #GLMs
              "mgcv") #GAMs 
lapply(packages, library, character.only = TRUE)

load("tramacastilla_disturbance_palaeoecology.RData")
```

# 1. Richness with rarefaction
## 1.1. Palynological richness with rarefaction
Function rarefy gives the expected richness in random subsamples of size "sample" from the community.
The size of "sample" should be smaller than total community size.
If "sample" is a vector, rarefaction of all observations is performed for each sample size separately

```{r}
tram20_pollen_holocene_richness <- tram20_pollen_holocene %>%
  dplyr::select(-c(SampleId, age, depth, `Terrestrial pollen sum`:Lycopodium, `Evergreen trees`:Trees, Shrubs, Herbs:`Charcoal <150`))

S <- specnumber(tram20_pollen_holocene_richness) # observed number of species
raremax <- min(rowSums(tram20_pollen_holocene_richness))
Srare <- rarefy(tram20_pollen_holocene_richness, raremax)
plot(S, Srare, xlab = "Observed No. of Species", ylab = "Rarefied No. of Species")
abline(0, 1)

rarecurve(tram20_pollen_holocene_richness, step = 20, sample = raremax, col = "blue", cex = 0.6)
```

Create dataframe with rarefaction richness as a variable
```{r}
tram20_pollen_holocene_rarefaction <- tram20_pollen_holocene %>%
  cbind(Srare) %>% #add rarefaction as a new column
  rename(Rarefaction = Srare) %>%
  relocate(Rarefaction, .after = "No of taxa")
```

# 2. Evenness

Create a function that allows for the estimation of different diversity indexes
```{r}
calculate_evenness <- function(abundances, method = "pielou") {
  S <- specnumber(abundances)
  if (method == "pielou") {
    H <- diversity(abundances, "shannon")
    J <- H / log(S)
    return(J)
  } else if (method == "simpson") {
    lambda <- 1 - diversity(abundances, "simpson")
    N2 <- 1 / lambda
    E <- N2 / S
    return(E)
  }
}
```

## 2.1. Palynological evenness
Calculate Simpson index and plot
```{r}
evenness_simpson <- calculate_evenness(tram20_PAR %>%
                     dplyr::select(-c(depth, age)), "simpson")

tram_pollen_evenness_simpson <- data.frame(tram20_PAR %>% dplyr::select(age),
                                           evenness_simpson)

tram_pollen_evenness_simpson %>%
  ggplot(aes(x = age/1000, y = evenness_simpson)) +
  geom_line() +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  scale_y_continuous(limits = c(0.1, 0.4),
                     breaks = seq(0.1, 0.4, 0.2)) +
  theme_classic() +
  labs(x = "Age (ka BP)",
       y = "Evenness") +
  geom_point(data = RAI_animals_domestic %>%
               mutate(presence = ifelse(Cattle|Sheep|Goat > 0.0, 0.35, 0)), # orange points when any domestic animal is identified
             aes(x = age/1000, y = presence), color = "orange", size = 3) +
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_text(size = 11))
```

## 2.2. sedaDNA evenness ######
Calculate Simpson index and plot
```{r}
evenness_simpson_RAI <- calculate_evenness(RAI_plants %>%
                                         dplyr::select(-c(sample_id, depth, age)), "simpson")

tram_RAI_evenness_simpson <- data.frame(RAI_plants %>% dplyr::select(age),
                                        evenness_simpson_RAI)

tram_RAI_evenness_simpson %>%
  ggplot(aes(x = age/1000, y = evenness_simpson_RAI)) +
  geom_line() +
  scale_x_reverse(limits = c(12, 0.5),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  scale_y_continuous(limits = c(0.1, 0.4),
                     breaks = seq(0.1, 0.4, 0.2)) +
  theme_classic() +
  labs(x = "Age (ka BP)",
       y = "Evenness") +
  geom_point(data = RAI_animals_domestic %>%
               mutate(presence = ifelse(Cattle|Sheep|Goat > 0.0, 0.35, 0)), # orange points when any domestic animal is identified
             aes(x = age/1000, y = presence), color = "orange", size = 3) +
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_text(size = 11))
```

# 3. Rate of Change

## 3.1. Rate of Change of PAR

### 3.1.1. Import data for Rate of Change

```{r}
tram20_RoC_PAR <- tram20_PAR %>%
  left_join(tram20_depth_ages_sed_rate %>%
              dplyr::select(SampleId, age), by = "age") %>%
  relocate(SampleId, .before = depth) %>%
  dplyr::select(-c(depth, age, Cyperaceae:Undetermined)) %>% #keep only terrestrial pollen types
  rename("sample_id" = "SampleId")

tram20_RoC_PAR %>%
  mutate(sum = rowSums(across(where(is.numeric)))) %>%
  dplyr::select(sample_id, sum) %>%
  filter(sum == min(sum)) #307
```

### 3.1.2. BChron age depth model
```{r}
head(tram20_BChron_data)
Tram20Chron = with(tram20_BChron_data, Bchronology(ages=ages,
                                                   ageSds=ageSds, 
                                                   calCurves=calCurves,
                                                   positions=position, 
                                                   positionThicknesses=thickness,
                                                   ids=id, 
                                                   predictPositions=seq(0,1500,by=10)))
plot(Tram20Chron)
```

Age uncertainties from BChron age-depth model
```{r}
head(tram20_level) # file with sample_id and depths associated

age_position <- Bchron:::predict.BchronologyRun(object = Tram20Chron, newPositions = tram20_level$depth)

age_uncertainties <- age_position %>%  # assign ages to each depth
  as.data.frame() %>%
  mutate_all(., as.integer) %>%
  as.matrix()
colnames(age_uncertainties) <- tram20_level$sample_id
```

```{r}
tram20_level_predicted <- tram20_level %>%
  mutate(age = apply(age_uncertainties, 2, stats::quantile,  probs = 0.5))

head(tram20_level_predicted)

age.RoC.PAR <- tram20_depth_ages %>%
  rename("age" = "median") %>%
  dplyr::select(SampleId, age) %>%
  mutate(age = as.numeric(age)) %>%
  rename("sample_id" = "SampleId") %>%
  filter(sample_id %in% tram20_RoC_PAR$sample_id)
```

### 3.1.3. Perform Rate of Change

Scenario 4 - Estimating RoC for each level subsampling data and calculating age uncertainties
```{r}
set_randomisations <- 1000
scenario_4_PAR <- RRatepol::estimate_roc(data_source_community = tram20_RoC_PAR,
                                              data_source_age = age.RoC.PAR,
                                              dissimilarity_coefficient = "chisq",
                                              working_units = "levels",
                                              time_standardisation = 500,
                                              smooth_method = "shep",
                                              standardise = TRUE,
                                              n_individuals = 29, # set the number of PAR
                                              rand = set_randomisations,
                                              use_parallel = TRUE,
                                              age_uncertainty = age_uncertainties) # Add the uncertainty matrix
RRatepol::plot_roc(data_source = scenario_4_PAR)
```

Scenario 4b - Estimating RoC detecting peak points
```{r}
scenario_4_PAR_peak <- RRatepol::detect_peak_points(data_source = scenario_4_PAR,
                                                         sel_method = "trend_non_linear")
scenario_4_PAR_peak$Peak <- as.vector(scenario_4_PAR_peak$Peak)

scenario_4_PAR_peak <- scenario_4_PAR_peak %>%
  mutate(ROC_peak = case_when (Peak == "TRUE" ~ ROC,
                               Peak == "FALSE" ~ NA))

RRatepol::plot_roc(data_source = scenario_4_PAR_peak, peaks = TRUE)
```

Plot
```{r}
scenario_4_PAR_peak %>%
  ggplot(aes(x = Age/1000))+
  theme_classic() +
  geom_ribbon(aes(ymin = ROC_dw, ymax = ROC_up), fill = "grey") +
  geom_line(aes(y = ROC)) + 
  geom_point(size = 3, aes(y = ROC_peak), shape = 21, fill = "green") +
  scale_x_reverse(limits = c(max(CHAR$age)/1000, min(CHAR$age)/1000),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  xlab("Age (ka BP)") + 
  ylab("Rate of Change (RoC) from PAR")
```

## 3.2. Rate of Change with SAR of coprophilous

### 3.2.1. Import data for Rate of Change

```{r}
tram20_RoC_SAR <- tram20_SAR_coprophilous %>%
  dplyr::select(-age) %>% #exclude age
  left_join(tram20_level, by = "depth") %>%
  dplyr::select(-depth) %>%
  relocate(sample_id, .before = 0)

tram20_RoC_SAR %>%
  mutate(sum = rowSums(across(where(is.numeric)))) %>%
  dplyr::select(sample_id, sum) %>%
  filter(sum > 0) %>%
  filter(sum == min(sum)) #2.57

age.RoC.SAR <- age.RoC.PAR %>%
  filter(age %in% tram20_SAR$age) %>%
  dplyr::select(sample_id, age)

age_uncertainties_SAR <- age_uncertainties %>%
  as.data.frame() %>%
  dplyr::select(any_of(age.RoC.SAR$sample_id)) %>%
  as.matrix()
```

### 3.2.2. Perform Rate of Change

Scenario 4 - Estimating RoC for each level subsampling data and calculating age uncertainties
```{r}
set_randomisations <- 1000
scenario_4_SAR <- RRatepol::estimate_roc(data_source_community = tram20_RoC_SAR,
                                         data_source_age = age.RoC.SAR,
                                         dissimilarity_coefficient = "chisq",
                                         working_units = "levels",
                                         time_standardisation = 500,
                                         smooth_method = "shep",
                                         standardise = TRUE,
                                         n_individuals = 3, # set the number of SAR
                                         rand = set_randomisations,
                                         use_parallel = TRUE,
                                         age_uncertainty = age_uncertainties_SAR) # Add the uncertainty matrix
RRatepol::plot_roc(data_source = scenario_4_SAR)
```

Scenario 4b - Estimating RoC detecting peak points
```{r}
# first, solve problems associated to the presence of 0 in the dataframe
scenario_4_SAR_clean <- scenario_4_SAR %>%
  mutate(across(where(is.numeric), ~ replace(., . == 0, 0.00001)))

scenario_4_SAR_peak <- RRatepol::detect_peak_points(data_source = scenario_4_SAR_clean,
                                                    sel_method = "trend_non_linear")

scenario_4_SAR_peak$Peak <- as.vector(scenario_4_SAR_peak$Peak)

scenario_4_SAR_peak <- scenario_4_SAR_peak %>%
  mutate(ROC_peak = case_when (Peak == "TRUE" ~ ROC,
                               Peak == "FALSE" ~ NA))

RRatepol::plot_roc(data_source = scenario_4_SAR_peak, peaks = TRUE)
```

Plot
```{r}
scenario_4_SAR_peak %>%
  ggplot(aes(x = Age/1000))+
  theme_classic() +
  geom_ribbon(aes(ymin = ROC_dw, ymax = ROC_up), fill = "grey") +
  geom_line(aes(y = ROC)) + 
  geom_point(size = 3, aes(y = ROC_peak), shape = 21, fill = "green") +
  scale_x_reverse(limits = c(max(CHAR$age)/1000, min(CHAR$age)/1000),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), "")) +
  xlab("Age (ka BP)") + 
  ylab("Rate of Change (RoC) from SAR")
```

# 4. Changepoint analysis with CHAR data

```{r}
tram20_CHAR_cpt <- CHAR %>%
  dplyr::select(age, CHAR.tram) %>%
  arrange(age)

out=envcpt(tram20_CHAR_cpt$CHAR.tram) # <- The key instruction!
plot(out,type='fit') # plots the fits
plot(out,type="aic") # plots the aic values
plot(out,type="bic") # plots the bic values
which.min(AIC(out)); which.min(BIC(out))

model_CHAR <- envcpt(tram20_CHAR_cpt$CHAR.tram, models = "meanar2cpt") #here the chosen model
summary(model_CHAR$meanar2cpt)
plot(model_CHAR$meanar2cpt)
```

Assign ages to each changepoint and extract positions from changepoints
```{r}
cpts_CHAR <- cpts(model_CHAR$meanar2cpt) # rows where changepoints are
print(cpts_CHAR)

coef_ar_CHAR <- model_CHAR$meanar2cpt@param.est$ar

changepoint_ages_CHAR <- tram20_CHAR_cpt$age[cpts_CHAR] # ages to each row: 1830 2003 6145
```

Plot
```{r}
tram20_CHAR_cpt %>%
  ggplot(aes(x = age/1000, y = CHAR.tram)) +
  geom_line() +
  geom_vline(xintercept = changepoint_ages_CHAR/1000, color = "red", linetype = "dashed") +
  labs(x = "Age (ka BP)", y = "CHAR (µm2·cm-2·yr-1)", title = "Changepoints using CHAR") +
  theme_classic() +
  scale_x_reverse(limits = c(max(CHAR$age)/1000, min(CHAR$age)/1000),
                  breaks = seq(0.5, 12, by = 0.5),
                  labels = ifelse(seq(0.5, 12, by = 0.5) %% 1 == 0, as.character(seq(0.5, 12, by = 0.5)), ""))
```

# 5. Cross-correlations for different periods: before and after 6.5 ka BP

Import data
```{r}
char_full <- CHAR %>%
  filter(age %in% tram20_pollen_holocene$age)
```

Create a function to perform cross-correlations for both periods
```{r}
tram_ccf <- function(data, taxa, char_full_data, period = "all", 
                     age_threshold = 6500, lag_max = 5) {
  
  # Filter data based on period
  filtered_data <- switch(period,
                          "all" = data,
                          "before" = data %>% filter(age > age_threshold),
                          "after" = data %>% filter(age <= age_threshold))
  
  # Filter CHAR data to match the same period
  char_filtered <- char_full_data %>%
    filter(age %in% filtered_data$age) %>%
    dplyr::select(CHAR.tram) %>%
    pull()
  
  # Transform taxa data into temporal series
  taxa_ts <- filtered_data %>%
    dplyr::select({{taxa}}) %>%
    pull()
  
  # Detrending lineal: adjust simple linear model and extract residuals
  taxa_det <- resid(lm(taxa_ts ~ seq_along(taxa_ts)))
  char_det <- resid(lm(char_filtered ~ seq_along(char_filtered)))
  
  # Cross-correlation
  ccf_result <- ccf(taxa_det, char_det,
                    lag.max = lag_max,
                    plot = FALSE)
  
  ccf_df <- with(ccf_result, data.frame(lag, acf))
  
  n_obs <- length(taxa_det) # Significance
  ci <- qnorm((1 + 0.95) / 2) / sqrt(n_obs)  # 95%
  
  # Plot title with period info
  period_title <- switch(period,
                         "all" = "All Period",
                         "before" = paste("age >", age_threshold),
                         "after" = paste("age =<", age_threshold))
  
  # Plot
  if (period == "before") {
    p <- ggplot(ccf_df, aes(x = lag, y = acf)) +
      geom_bar(stat = "identity", width = 1,
               colour = "#666666", fill = "grey") +
      geom_hline(yintercept = c(-ci, ci),
                 linetype = "dashed", color = "red") +
      labs(y = paste(deparse(substitute(taxa))),
           x = "Lags") +
      theme_classic() +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#666666") +
      theme(axis.title.x = element_blank(),
            axis.title.y = element_text(face = "italic", size = 12, angle = 0,vjust = 0.5),
            axis.text.x = element_text(size = 11),
            axis.text.y = element_text(size = 11))
    
  } else if (period == "after") {
    p <- ggplot(ccf_df, aes(x = lag, y = acf)) +
      geom_bar(stat = "identity", width = 1,
               colour = "#666666", fill = "grey") +
      geom_hline(yintercept = c(-ci, ci),
                 linetype = "dashed", color = "red") +
      theme_classic() +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#666666") +
      theme(axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.x = element_text(size = 11),
            axis.text.y = element_text(size = 11))
  }
  
  return(list(plot = p, ccf_df = ccf_df, ci = ci, n = n_obs, period = period))
}
```

## 5.1. Cross-correlations for each taxa and periods

Abies
```{r}
abies_ccf_before <- tram_ccf(tram20_PAR, Abies, char_full, period = "before")
abies_ccf_after <- tram_ccf(tram20_PAR, Abies, char_full, period = "after")
```

Pinus
```{r}
pinus_ccf_before <- tram_ccf(tram20_PAR, Pinus, char_full, period = "before")
pinus_ccf_after <- tram_ccf(tram20_PAR, Pinus, char_full, period = "after")
```

Fagus
```{r}
fagus_ccf_before <- tram_ccf(tram20_PAR, Fagus, char_full, period = "before")
fagus_ccf_after <- tram_ccf(tram20_PAR, Fagus, char_full, period = "after")
```

Betula
```{r}
betula_ccf_before <- tram_ccf(tram20_PAR, Betula, char_full, period = "before")
betula_ccf_after <- tram_ccf(tram20_PAR, Betula, char_full, period = "after")
```

Corylus
```{r}
corylus_ccf_before <- tram_ccf(tram20_PAR, Corylus, char_full, period = "before")
corylus_ccf_after <- tram_ccf(tram20_PAR, Corylus, char_full, period = "after")
```

Deciduous Quercus
```{r}
decquercus_ccf_before <- tram_ccf(tram20_PAR %>%
                                    rename("Dec. Quercus" = "Deciduous Quercus"),
                                  `Dec. Quercus`, char_full, period = "before")
decquercus_ccf_after <- tram_ccf(tram20_PAR %>%
                                   rename("Dec. Quercus" = "Deciduous Quercus"),
                                 `Dec. Quercus`, char_full, period = "after")
```

Poaceae
```{r}
poaceae_ccf_before <- tram_ccf(tram20_PAR, Poaceae, char_full, period = "before")
poaceae_ccf_after <- tram_ccf(tram20_PAR, Poaceae, char_full, period = "after")
```

Palynological richness
```{r}
richness_ccf_before <- tram_ccf(tram20_pollen_holocene_rarefaction %>%
                                  rename("Richness" = "Rarefaction"),
                                Richness, char_full, period = "before")
richness_ccf_after <- tram_ccf(tram20_pollen_holocene_rarefaction %>%
                                 rename("Richness" = "Rarefaction"),
                               Richness, char_full, period = "after")
```

Palynological evenness
```{r}
evenness_ccf_before <- tram_ccf(tram_pollen_evenness_simpson%>%
                                  rename("Evenness" = "evenness_simpson"),
                                Evenness, char_full, period = "before")
evenness_ccf_after <- tram_ccf(tram_pollen_evenness_simpson%>%
                                 rename("Evenness" = "evenness_simpson"),
                               Evenness, char_full, period = "after")
```

## 5.2. Create plots of cross-correlations

```{r}
p.abies.ccf_before <- abies_ccf_before$plot
p.abies.ccf_after <- abies_ccf_after$plot

p.pinus.ccf_before <- pinus_ccf_before$plot
p.pinus.ccf_after <- pinus_ccf_after$plot

p.fagus.ccf_before <- fagus_ccf_before$plot
p.fagus.ccf_after <- fagus_ccf_after$plot

p.betula.ccf_before <- betula_ccf_before$plot
p.betula.ccf_after <- betula_ccf_after$plot

p.corylus.ccf_before <- corylus_ccf_before$plot
p.corylus.ccf_after <- corylus_ccf_after$plot

p.decquercus.ccf_before <- decquercus_ccf_before$plot
p.decquercus.ccf_after <- decquercus_ccf_after$plot

p.poaceae.ccf_before <- poaceae_ccf_before$plot
p.poaceae.ccf_after <- poaceae_ccf_after$plot

p.richness.ccf_before <- richness_ccf_before$plot
p.richness.ccf_after <- richness_ccf_after$plot

p.evenness.ccf_before <- evenness_ccf_before$plot
p.evenness.ccf_after <- evenness_ccf_after$plot
```

# 6. Structural Equation Modelling (SEM)

Prepare all data
```{r}
RAI_plants_richness <- RAI_plants %>%
  mutate(richness = rowSums(across(c(`Abies alba`:Ziziphora), ~ .x > 0), na.rm = TRUE))

RAI_plants_richness_binned <- RAI_plants_richness %>%
  mutate(binned_age = sapply(age, function(x) bins_pollen[which.min(abs(bins_pollen-x))])) %>%
  mutate(RAI_Fagus = Fagus + `Fagus sylvatica`) %>%
  group_by(binned_age) %>%
  summarise(richness = mean(richness, na.rm = TRUE),
            RAI_Fagus = mean(RAI_Fagus, na.rm = TRUE)) %>%
  rename("age" = "binned_age",
         "plant_richness_dna" = "richness") %>%
  ungroup()

head(RAI_plants_groups)

RAI_plants_groups_binned <- RAI_plants_groups %>%
  mutate(binned_age = sapply(age, function(x) bins_pollen[which.min(abs(bins_pollen-x))])) %>%
  group_by(binned_age) %>%
  summarise(across(c(`Abies alba`, `Open landscape`, `Novel herbaceous taxa`),
                   ~ mean(., na.rm = TRUE))) %>%
  rename("age" = "binned_age",
         "RAI_Abies" = "Abies alba",
         "RAI_Open" = "Open landscape",
         "RAI_Novel" = "Novel herbaceous taxa") %>%
  ungroup()

tram_RAI_evenness_simpson_binned <- tram_RAI_evenness_simpson %>%
  mutate(binned_age = sapply(age, function(x) bins_pollen[which.min(abs(bins_pollen-x))])) %>%
  group_by(binned_age) %>%
  summarise(evenness_simpson_RAI  = mean(evenness_simpson_RAI, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup()

head(Mendukilo)

Mendukilo_binned <- Mendukilo %>%
  mutate(binned_age = sapply(age, function(x) bins_pollen[which.min(abs(bins_pollen-x))])) %>%
  group_by(binned_age) %>%
  summarise(d13C.Mendukilo = mean(d13C.Mendukilo, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()
```

All data in one dataframe
```{r}
data_SEM <- RAI_animals_domestic_binned %>%
  left_join(RAI_plants_richness_binned, by = "age") %>%
  left_join(CHAR %>%
              dplyr::select(CHAR.tram, age), by = "age") %>%
  left_join(tram_pollen_evenness_simpson, by = "age") %>%
  left_join(tram_RAI_evenness_simpson_binned, by = "age") %>%
  left_join(tram20_SAR %>%
              dplyr::select(age, `Facultative coprophilous`, `Obligate coprophilous`), by = "age") %>%
  rename("fac_coprophilous" = "Facultative coprophilous",
         "ob_coprophilous" = "Obligate coprophilous") %>%
  left_join(Mendukilo_binned, by = "age") %>%
  left_join(RAI_plants_groups_binned, by = "age") %>%
  left_join(tram20_PAR %>%
              dplyr::select(age, Abies, Fagus, Poaceae) %>%
              rename("PAR_Abies" = "Abies",
                     "PAR_Fagus" = "Fagus",
                     "PAR_Poaceae" = "Poaceae"), by = "age")
```

Transform variables as PAR and RAI have many 0s, and SEM assumes multivariate normality, so log-transformations improve normality, lineality and homocedasticity
```{r}
data_transformed <- data_SEM %>%
  mutate(log_RAI_Abies = log10(RAI_Abies + 1e-7), # add small values for 0
         log_RAI_Fagus = log10(RAI_Fagus + 1e-7),
         log_RAI_Open = log10(RAI_Open + 1e-7),
         log_RAI_Novel = log10(RAI_Novel + 1e-7),
         log_PAR_Abies = log10(PAR_Abies + 1),
         log_PAR_Fagus = log10(PAR_Fagus + 1),
         log_PAR_Poaceae = log10(PAR_Poaceae + 1),
         log_CHAR = log10(CHAR.tram + 1),
         log_copro = log10(ob_coprophilous + 1),
         # standardise (optional, helps interpretation)
         z_d13C = as.numeric(scale(d13C.Mendukilo)),
         z_Cattle = as.numeric(scale(Cattle)),
         z_Sheep = as.numeric(scale(Sheep)),
         z_Goat = as.numeric(scale(Goat)),
         z_richness_RAI = as.numeric(scale(plant_richness_dna)),
         z_evenness_polen = as.numeric(scale(evenness_simpson)),
         z_evenness_RAI = as.numeric(scale(evenness_simpson_RAI)))
```

Explore correlations between variables
```{r}
vars_cor <- data_transformed %>% 
  dplyr::select(log_RAI_Fagus, log_RAI_Abies, Cattle, Sheep, Goat, 
                log_copro, log_CHAR, z_d13C, log_RAI_Open)

cor_matrix <- cor(vars_cor, use = "complete.obs")
corrplot::corrplot(cor_matrix, method = "color", 
                   type = "upper", tl.col = "black")
```

## 6.1. Prepare data for communities
```{r}
data_SEM_communities <- data_transformed %>%
  dplyr::select(age, z_Cattle, z_Sheep, z_Goat, log_CHAR, z_richness_RAI, z_evenness_polen,
                z_evenness_RAI, log_copro, z_d13C) %>%
  left_join(tram20_PAR_groups_communities %>%
              rename_with(~ gsub("^", "PAR_", .x), .cols = Abies_Fagus:`temperate forest`),
            by = "age") %>%
  mutate(log_PAR_Abies_Fagus = log10(PAR_Abies_Fagus + 1), # add small values for 0
         log_PAR_grassland = log10(PAR_grassland + 1),
         log_PAR_mediterranean = log10(PAR_mediterranean + 1),
         log_PAR_pinewoods  = log10(PAR_pinewoods + 1),
         log_PAR_riverine = log10(`PAR_riverine community` + 1),
         log_PAR_rocky = log10(`PAR_rocky community` + 1),
         log_PAR_shrubland  = log10(PAR_shrubland  + 1),
         log_PAR_temperate = log10(`PAR_temperate forest` + 1),
         log_PAR_pastureland = log10(PAR_pastureland + 1))
```

Explore correlations between variables
```{r}
cor_matrix_communities <- cor(data_SEM_communities %>% dplyr::select(-age, -depth), use = "complete.obs")
corrplot::corrplot(cor_matrix_communities, method = "color", 
                   type = "upper", tl.col = "black")
```

## 6.2. Model with selected variables

```{r echo=T, results='hide'}
tram_SEM_PAR_communities_selected <-  '# Latent variables
    Herbivory =~ NA*z_Cattle + z_Sheep + log_copro
    Herbivory ~~ 1*Herbivory
    
    # Structural regressions
    # Effects of temperature (d13C) on disturbances
    Herbivory ~ z_d13C
    
    # Effects of disturbances on communities
    log_PAR_Abies_Fagus ~ Herbivory
    log_PAR_grassland ~ log_CHAR + z_d13C
    log_PAR_pinewoods ~ log_CHAR
    log_PAR_temperate ~ log_CHAR + Herbivory
    log_PAR_mediterranean ~ z_d13C
    
    #Effects of disturbances on heterogeneity
    z_evenness_polen ~ Herbivory + log_CHAR
    
    # Residual covariance between disturbances
    Herbivory ~~ log_CHAR
'
fit_no_normal_8 <- cfa(tram_SEM_PAR_communities_selected, data = data_SEM_communities, test = "bootstrap", verbose = TRUE, bootstrap = 1000)
summary(fit_no_normal_8, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
fitMeasures(fit_no_normal_8, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "srmr"))
```

See standardised effects (comparable)
```{r}
parameters_signif_nonormal8 <- parameterEstimates(fit_no_normal_8, standardized = TRUE) %>%
  filter(op == "~") %>%
  mutate(variable = paste(lhs, "←", rhs),
         dir = ifelse(std.all > 0, "positive", "negative")) %>%
  dplyr::select(variable, est, std.all, dir, pvalue) %>%
  arrange(desc(abs(std.all)))
```

Figure of standardised coefficients
```{r}
parameters_signif_nonormal8 %>%
  filter(pvalue < 0.05) %>%
  ggplot(aes(x = reorder(variable, std.all), y = std.all)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Relation", y = "Standardised coefficient",
       title = "Significative effects on SEM") +
  theme_minimal()
```

Interpretation:
Adequacy of the model -> NEARLY ADECUATE
   p-value = 0.463
   CFI = 1
   robust RMSEA = 0.002
   SRMS = 0.077
Significative effects on Abies_Fagus community:
   Herbivory (+)
Significative effects on herbivory:
   temperature (+)
Significative effects on grasslands:
   temperature (+)
   fire (+)
Significative effects on pinewoods:
   fire (+)
Significative effects on temperate forest:
   herbivory (-)
Significative effects on pollen evenness:
   herbivory (+)
   fire (-)

# 7. GAMs -- Turnover vegetation drivers

## 7.1 Prepare data with all variables

```{r}
bins_drivers <- scenario_4_RAI_peak$Age

Mendukilo_binned_drivers <- Mendukilo %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(d13C.Mendukilo = mean(d13C.Mendukilo, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

CHAR_binned_drivers <- CHAR %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(CHAR.tram = mean(CHAR.tram, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

RAI_animals_domestic_binned_drivers <- RAI_animals_domestic %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(Cattle = mean(Cattle, na.rm = TRUE),
            Sheep = mean(Sheep, na.rm = TRUE),
            Goat = mean(Goat, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

tram20_copro_binned_drivers <- tram20_SAR %>%
  dplyr::select(age, `Facultative coprophilous`, `Obligate coprophilous`) %>%
  rename("fac_coprophilous" = "Facultative coprophilous",
         "ob_coprophilous" = "Obligate coprophilous") %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(fac_coprophilous = mean(fac_coprophilous, na.rm = TRUE),
            ob_coprophilous = mean(ob_coprophilous, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

tram20_PAR_sum_binned_drivers <-tram20_PAR_sum %>%
  dplyr::select(age, sum) %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(PAR_sum = mean(sum, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

scenario_4_PAR_peak_binned <- scenario_4_PAR_peak %>%
  dplyr::select(Age, ROC) %>%
  rename("age" = "Age") %>%
  mutate(binned_age = sapply(age, function(x) bins_drivers[which.min(abs(bins_drivers-x))])) %>%
  group_by(binned_age) %>%
  summarise(ROC = mean(ROC, na.rm = TRUE)) %>%
  rename("age" = "binned_age") %>%
  ungroup() %>%
  as.data.frame()

data_drivers_turnover <- scenario_4_RAI_peak %>%
  dplyr::select(Age, ROC) %>%
  rename("age" = "Age",
         "ROC_RAI" = "ROC") %>%
  left_join(scenario_4_PAR_peak_binned %>%
              rename("ROC_PAR" = "ROC"), by = "age") %>%
  left_join(Mendukilo_binned_drivers, by = "age") %>%
  left_join(CHAR_binned_drivers, by = "age") %>%
  left_join(RAI_animals_domestic_binned_drivers, by = "age") %>%
  left_join(tram20_copro_binned_drivers, by = "age") %>%
  left_join(tram20_PAR_sum_binned_drivers, by = "age")

data_drivers_turnover_transformed <- data_drivers_turnover %>%
  mutate(z_ROC_RAI = as.numeric(scale(ROC_RAI)),
         z_ROC_PAR = as.numeric(scale(ROC_PAR)),
         z_d13C = as.numeric(scale(d13C.Mendukilo)),
         z_Cattle = as.numeric(scale(Cattle)),
         z_Sheep = as.numeric(scale(Sheep)),
         z_Goat = as.numeric(scale(Goat)),
         z_CHAR = as.numeric(CHAR.tram),
         log_ob_copro = log10(ob_coprophilous + 1),
         log_fac_copro = log10(fac_coprophilous + 1),
         log_PAR = log10(PAR_sum + 1)) %>%
  drop_na()
```

## 7.2. Individual GAMs

See the effect of temperature on Rate of Change of sedaDNA
```{r}
gam_drivers_1 <- gam(z_ROC_RAI ~ z_d13C, data = data_drivers_turnover_transformed)
summary_gam1 <- summary(gam_drivers_1) #R2 = 41% and significative
```

See the effect of temperature on Rate of Change of pollen
```{r}
gam_drivers_1_PAR <- gam(z_ROC_PAR ~ z_d13C, data = data_drivers_turnover_transformed)
summary_gam1_PAR <- summary(gam_drivers_1_PAR) #R2 = 45% and significative
```

See the effect of herbivory (cattle) on Rate of Change of sedaDNA
```{r}
gam_drivers_2 <- gam(z_ROC_RAI ~ z_Cattle, data = data_drivers_turnover_transformed)
summary_gam2 <- summary(gam_drivers_2) #R2 = 25% and significative
```

See the effect of herbivory (cattle) on Rate of Change of pollen
```{r}
gam_drivers_2_PAR <- gam(z_ROC_PAR ~ z_Cattle, data = data_drivers_turnover_transformed)
summary_gam2_PAR <- summary(gam_drivers_2_PAR) #R2 = 33% and significative
```

See the effect of herbivory (Sheep) on Rate of Change of sedaDNA
```{r}
gam_drivers_3 <- gam(z_ROC_RAI ~ z_Sheep, data = data_drivers_turnover_transformed)
summary_gam3 <- summary(gam_drivers_3) #R2 = 1% and not significative
```

See the effect of herbivory (Sheep) on Rate of Change of pollen
```{r}
gam_drivers_3_PAR <- gam(z_ROC_PAR ~ z_Sheep, data = data_drivers_turnover_transformed)
summary_gam3_PAR <- summary(gam_drivers_3_PAR) #R2 = 1% and not significative
```

See the effect of herbivory (coprophilous fungi) on Rate of Change of sedaDNA
```{r}
gam_drivers_4 <- gam(z_ROC_RAI ~ log_ob_copro, data = data_drivers_turnover_transformed)
summary_gam4 <- summary(gam_drivers_4) #R2 = 47% and significative
```

See the effect of herbivory (coprophilous fungi) on Rate of Change of pollen
```{r}
gam_drivers_4_PAR <- gam(z_ROC_PAR ~ log_ob_copro, data = data_drivers_turnover_transformed)
summary_gam4_PAR <- summary(gam_drivers_4_PAR) #R2 = 25% and significative
```

See the effect of fire on Rate of Change of sedaDNA
```{r}
gam_drivers_5 <- gam(z_ROC_RAI ~ z_CHAR, data = data_drivers_turnover_transformed)
summary_gam5 <- summary(gam_drivers_5) #R2 = 0.05% and not significative
```

See the effect of fire on Rate of Change of pollen
```{r}
gam_drivers_5_PAR <- gam(z_ROC_PAR ~ z_CHAR, data = data_drivers_turnover_transformed)
summary_gam5_PAR <- summary(gam_drivers_5_PAR) #R2 = 0.2% and not significative
```